---
title: <span style="color:orange">Managing</span> <span style="color:#FFFFFF">R Programmers in Pharma</span>

author: <span style="color:#FFFFFF">twitter:@rinpharma github:philbowsher</span>
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [ "include/rstudio.css" ]
    nature:
      countIncrementalSlides: yes
      highlightLines: yes
      highlightStyle: github
      ratio: 16:9
resource_files:
  - include
  - images
---


```{r, include=FALSE}
rmarkdown::output_metadata$set(
  rsc_output_files = list.files("images", recursive = TRUE, full.names = TRUE)
)
```


class: subtitle-slide, white, middle, center

# History

---

class: panel-narrow-slide, left

# R
- https://statfr.blogspot.com/2018/08/r-generation-story-of-statistical.html
- https://cran.r-project.org/
- https://github.com/tidyverse/ggplot2
- http://hadley.nz/
- https://www.nytimes.com/2009/01/07/technology/business-computing/07program.html?pagewanted=2
- http://appliedpredictivemodeling.com/blog
- https://shiny.rstudio.com/
---

class: panel-narrow-slide, left

# FDA
- https://channel9.msdn.com/Events/useR-international-R-User-conference/useR2016/Using-R-in-a-regulatory-environment-FDA-experiences
- http://washstat.org/presentations/20181024/Schuette.pdf
- https://www.cdisc.org/system/files/all/event/restricted/2019_Japan/16_AShapiro%2520CDISC%2520Japan%25202019%2520Presentation.pdf
- https://www.accessdata.fda.gov/drugsatfda_docs/nda/2016/208573Orig1s000ClinPharmR.pdf
- https://www.accessdata.fda.gov/drugsatfda_docs/nda/2017/209296Orig1s000ClinPharmR.pdf
---

class: panel-narrow-slide, left

# Tools
- https://www.novartis.com/careers/career-search/job-details/295774BR
- https://rinterface.com/shiny/talks/zurichR2020/index.html#slide=2
- https://www.roche.com/careers/jobs/jobsearch/job.htm?id=E-202004-111441&locale=en&title=Data+Management+Engineer+II
- https://www.biospace.com/job/2087510/clinical-trial-statistical-programmer-/
---

class: panel-narrow-slide, left

# Tools Cont.
- https://boards.greenhouse.io/dnli/jobs/2132890
- https://lensa.com/data-scientistengineer-r-code-aws-big-data-jobs/san-francisco/jd/74536274507bc9adbc83babb7cc9dff9
- https://jobsvaultglobal.com/jobs/r-shiny-programmer-remote-opportunity-san-rafael-california/146494872-2/
---

class: panel-narrow-slide, left

# Working groups
- https://www.phuse.eu/blog/working-groups-2019-end-of-year-summary
- https://www.phuse.eu/standard-analyses-code-sharing
- https://www.phusewiki.org/wiki/index.php?title=Standard_Analyses_%26_Code_Sharing
- https://www.phusewiki.org/wiki/index.php?title=Best_Practices_for_Interactive_Analyses_for_Decision_Making_%26_Submissions
- https://www.phusewiki.org/wiki/index.php?title=Links_to_Resources_and_Knowledge
---


class: panel-narrow-slide, left

# Working groups...
- https://rinpharma.github.io/website2018/program/using-r-in-a-regulatory-environment-some-fda-perspectives.html
- https://www.meetup.com/Bay-Area-Biotech-Pharma-Statistics-Workshop-BBSW/events/265042071/
- https://www.pharmar.org/white-paper/
- https://github.com/rinpharma/rinpharma2019program
---

class: panel-narrow-slide, left

# Training
- https://www.psiweb.org/events/event-item/2020/06/11/default-calendar/psi-conference-webinar-psi-rss-prize-winners-session
- https://medicine.yale.edu/cbds/train_collab/
- RStudio Cloud
---

class: panel-narrow-slide, left

# Tidyverse Intro
- [Flatiron](https://rstudio.com/resources/rstudioconf-2020/flatironkitchen-how-we-overhauled-a-frankensteinian-sql-workflow-with-the-tidyverse/)
- https://www.pharmasug.org/proceedings/2019/DV/PharmaSUG-2019-DV-169.pdf
- https://www.linkedin.com/in/oleksandr-babych-777b00170/?originalSubdomain=ua
- https://www.enar.org/meetings/spring2020/program/Education_Program.pdf
- https://accp1.org/documents/annualmeeting/2019/2019ACCPProgram_Final.pdf
- https://www.ohsu.edu/octri/data-visualization-r-and-ggplot2
---


class: panel-narrow-slide, left

# Tidyverse
- https://www.bicycletherapeutics.com/wp-content/uploads/Oct232019_MT1MMP_TRIPLEposter_FINAL.pdf
- https://www.user2019.fr/talk_schedule/
- https://www.user2019.fr/static/pres/lt252425.pdf
- https://www.user2019.fr/static/pres/lt252371.pdf
- https://www.pharmasug.org/proceedings/2018/BB/PharmaSUG-2018-BB22.pdf
- http://www.mbswonline.com/upload/presentation_6-22-2018-14-25-55.introducing%20r_tidyverse%20to%20clinical%20statistical%20programming.pdf
---


class: subtitle-slide, white, middle, center

# What is the Tidyverse?

---



class: panel-narrow-slide, left

# Tidyverse is a...
- a package of packages that...
- share an underlying design philosophy, grammar, and data structures
- https://www.tidyverse.org/
---


class: panel-narrow-slide, left

# R for Data Science

- [Tools needed in a typical data science project](http://r4ds.had.co.nz/introduction.html)


---




class: panel-narrow-slide, left


# Remember dplyr single table verbs

- filter() extract cases
- arrange() reorder cases
- group_by() group cases
- select() extract variables
- mutate() create new variables
- summarise() summarise variables / create cases
---

class: panel-narrow-slide, left


# Remember tidyr single table verbs

- pivot_longer() Pivot data from wide to long
- pivot_wider() Pivot data from long to wide
- spread() Spread a key-value pair across multiple columns
- gather() Gather columns into key-value pairs
---

class: panel-narrow-slide, left


# The pipe operator %>%
- Passes result on left into first argument of function on right
- And then
- mutate(babynames, percent = prop * 100)
- babynames %>% mutate(percent = prop * 100)
---

class: panel-narrow-slide, left


# Tibble Example

```{r}
library(tibble)
#library("tidyverse")
storms <-tribble(
~storm, ~wind, ~pressure, ~date,
"Alberto", 110, 1007, "2000-08-12",
"Alex", 45, 1009, "1998-07-30",
"Allison", 65, 1005, "1995-06-04",
"Ana", 40, 1013, "1997-07-01",
"Arlene", 50, 1010, "1999-06-13",
"Arthur", 45, 1010, "1996-06-21"
)

# filter(storms, wind == 45)

```
---


class: panel-narrow-slide, left


# ggplot2

- [Plotting system for R](https://rpubs.com/hadley/ggplot2-toolbox)
- [Created by Hadley Wickham](http://ggplot2.org/)
- [Powerful graphics language for creating elegant and complex plots](http://www.cookbook-r.com/Graphs/)
- [ggplot2 is a graphing library for R. Use ggplot2 with Plotly to create interactive, online ggplot2 charts with D3.js.](https://plot.ly/ggplot2/)
- plotly

---

class: panel-narrow-slide, left

# Databases Landscape
- db.rstuidio.com
- http://db.rstudio.com/overview
- https://db.rstudio.com/best-practices/select-interface/
- DBI
- dplyr
- dbplyr 
- odbc 
- https://github.com/edgararuiz/db_best_practices_2018


---

class: panel-narrow-slide, left

# DB writequeries using dplyr syntax
- https://datascienceplus.com/working-with-databases-in-r/
- https://rviews.rstudio.com/2017/10/18/database-queries-with-r/

---

class: panel-narrow-slide, left

# Example
- https://github.com/rstudio/webinars/tree/master/72-introduction-to-tidy-statistics-with-r
- https://rstudio.cloud/project/381034

---


class: subtitle-slide, white, middle, center

# Live Demo

---



class: panel-narrow-slide, left

# Shiny
- https://www.phusewiki.org/docs/2019_SDE_Shanghai/Xuan%20Yang.pdf
- https://ryo-n7.github.io/2019-07-21-user2019-reflections/
- https://cran.r-project.org/web/packages/safetyGraphics/vignettes/shinyUserGuide.html
- https://www.efspi.org/Documents/Events/Events%202019/Reg%20Stats/Presentations/3_3_Falgreen.pdf
- https://www.lexjansen.com/css-us/2019/PP28.pdf
- https://www.bayes-pharma.org/wp-content/uploads/2019/06/Session-7-6-Anna-Pohlmann-REVISED.pdf
---

class: panel-narrow-slide, left

# Shiny Cont.
- https://www.lexjansen.com/phuse/2019/dv/DV07.pdf
- https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1008489
- https://github.com/AbbVie-ComputationalGenomics/genetic-evidence-approval
- https://www.pharmasug.org/proceedings/2019/AD/PharmaSUG-2019-AD-048.pdf
- https://ascpt.onlinelibrary.wiley.com/doi/10.1002/cpt.1732
- https://cran.r-project.org/web/packages/gsDesign/index.html
---

class: subtitle-slide, white, middle, center

# Live Demo

---


class: panel-narrow-slide, left

# RMD
- https://github.com/BBSW-org/BBSW2019/blob/master/Day%202%20(Nov%208)/3.%20Lightning%20talk%20II/2.%20dcivello-interactive_reports_in_r.pptx
- https://rstudio.com/resources/rstudioconf-2020/r-for-graphical-clinical-trial-reporting/
- https://www.themillerlab.io/post/flexdashboards/
- https://www.pharmasug.org/proceedings/2020/DV/PharmaSUG-2020-DV-164.pdf
---

class: subtitle-slide, white, middle, center

# Live Demo

---

class: panel-narrow-slide, left

# Future
- https://github.com/BBSW-org/R-meetup/blob/master/2019-11-06/Artificial%20Intelligence%20%26%20Analytical%20Innovation.pdf
- https://blogs.rstudio.com/ai/posts/2018-01-29-dl-for-cancer-immunotherapy/
---


class: panel-narrow-slide, left

# Validation
- https://environments.rstudio.com/validation.html
- https://cran.r-project.org/web/views/ClinicalTrials.html
- https://www.r-project.org/doc/R-FDA.pdf
- https://github.com/andyofsmeg/RValidation
---


class: panel-narrow-slide, left

# Collaboration

.pull-left[
  # [R in Pharma](https://rinpharma.com)
  ]
.pull-right[ 
  # [Pharma R PSI](https://pharmar.org)
]

---


class: panel-narrow-slide, left

# Clincal Tables
- gt

---


class: subtitle-slide, white, middle, center

# Appendix

---




class: panel-narrow-slide, left

# Clinical Examples
- https://therbootcamp.github.io/Intro2DataScience_2018Oct/_sessions/CaseStudies/Clinical_Data_Case_Study.html
- https://github.com/therbootcamp/Intro2DataScience_2018Oct
- https://github.com/bioinformatics-core-shared-training/r-intermediate/blob/master/2.dplyr-intro.Rmd

---
class: panel-narrow-slide, left

# Clinical Examples
- https://therbootcamp.github.io/Intro2DataScience_2018Oct/_sessions/Wrangling/Wrangling_practical.html
- https://therbootcamp.github.io/Erfurt_2018June/_sessions/D1S2_Wrangling/Wrangling_practical_answers.html
- https://therbootcamp.github.io/BaselRBootcamp_2018April/schedule

---

class: panel-narrow-slide, left

# Other Examples
- https://peerj.com/preprints/3163.pdf
- https://dpastoor.github.io/simplest_bayes/dataset-creation.html
- https://learnr.me/examples/ggplot_part1.html
- https://rpubs.com/ledongnhatnam/242506
---
class: panel-narrow-slide, left

# Clinical Examples
- http://r-addict.com/2016/11/21/Optimal-Cutpoint-maxstat.html
- http://www.rpubs.com/RedBeren/FL_hospital
- https://nycdatascience.com/blog/student-works/exploring-response-biomarkers-clinical-trial/

---

class: panel-narrow-slide, left

# And more examples : )
- https://rpubs.com/ledongnhatnam/261629
- https://datacarpentry.org/R-genomics/04-dplyr.html
- https://github.com/datacarpentry/R-genomics/


